sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-11-04 doc/paper-jss/inla-comparison.Rmd
captions for timing figures
Edits

```{r inla-mesh, echo = TRUE}
loc_xy <- as.matrix(d[, c("X", "Y"), drop = FALSE])
inla_mesh <- INLA::inla.mesh.create(loc_xy, refine = TRUE, cutoff = 10)
spde <- inla.spde2.pcmatern(
  mesh = inla_mesh, # an smdTMB mesh could be inserted here with `sdmTMB_mesh$mesh`
  prior.range = c(range_min, prior_prob),
  prior.sigma = c(sigma_max, prior_prob)
)
```

>>>>

```{r inla-mesh, echo = TRUE}
loc_xy <- as.matrix(d[, c("X", "Y"), drop = FALSE])
inla_mesh <- INLA::inla.mesh.create(loc_xy, refine = TRUE, cutoff = 10)
spde <- inla.spde2.pcmatern(
  mesh = inla_mesh, 
  prior.range = c(range_min, prior_prob),
  prior.sigma = c(sigma_max, prior_prob)
)
```

### 2022-11-04 doc/paper-jss/owls.Rmd
Edits

We can run additional optimization steps using the `run_extra_optimization()` function.

@@

And confirm that the gradients are smaller:

@@

View the confidence intervals on both fixed and random (and variance) parameter estimates using `tidy()`.

@@

@@

We can inspect randomized quantile residuals (Fig.\ \ref{fig:qqnorm-owls}), although this method has known issues [@thygesen2017a].

@@

More accurate residuals can be generated using \pkg{tmbstan}, but this a slower process.
These are randomized quantile residuals where the fixed effects are fixed at their MLE (maximum likelihoood estimate) values and the random effects are sampled with MCMC via \pkg{tmbstan}/\pkg{Stan}.
To calculate these, we have to use the version of our model fit without REML.
We would ideally sample with more MCMC samples than this and assess convergence by setting `print_stan_model = TRUE`.
For the purposes of this building this appendix quickly, we will not run this code.

```{r stan-resids-owls, eval=FALSE}
mcmc_resids <- residuals(fit_owl, type = "mle-mcmc", mcmc_iter = 101, mcmc_warmup = 100)
```

We can also test for zero-inflation using the `simulate.sdmTMB()` function.
If our data were zero-inflated or overdispersed, we might have needed to use a delta ("hurdle") model instead (e.g., `delta_truncated_nbinom2()` or possibly even `delta_gamma()` if we were willing to treat the counts as continuous.

@@

And we can more formally test this with functions from the \pkg{DHARMa} package.

@@

```{r plot-all-effects-owls, echo=FALSE, eval=TRUE, dpi=220, fig.cap="Predicted winter owl counts that incorporate all fixed effects and random effects.", fig.pos='h', fig.align='center'}
ggplot(data = p_proj) +
  geom_point(data = p, aes(X, Y, colour = exp(est)), size = 0.4, alpha = 0.5) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    trans = "sqrt", # makes variation in low numbers visible
    # can also trim the extreme high values
    limits = c(0, quantile(exp(p$est), 0.995)), na.value = "yellow",
    guide = guide_colourbar(title.position = "top")
  ) +
  labs(colour = "Predicted counts") +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

```{r plot-spatiotemporal-effects-owls, echo=FALSE, eval=TRUE, dpi=210, fig.cap="Spatiotemporal random effects.", fig.pos='h', fig.align='center'}
ggplot(data = filter(p_proj, year > 1978)) +
  geom_point(
    data = filter(p, year > 1978), aes(X, Y, colour = epsilon_st),
    size = 0.4, alpha = 0.5
  ) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_gradient2(guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

```{r plot-fixed-effects-owls, echo=FALSE, eval=TRUE, dpi=210, fig.cap="The colour of each year represents the non-random field based elements in this model.", fig.pos='h', fig.align='center'}
ggplot(data = filter(p_proj, year > 1978)) +
  geom_point(
    data = filter(p, year > 1978), aes(X, Y, colour = exp(est_non_rf)),
    size = 0.4, alpha = 0.5
  ) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    breaks = c(0.03, 0.05, 0.07),
    guide = guide_colourbar(title.position = "top")
  ) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

>>>>

We can run additional optimization steps with a Newton optimizer using the `run_extra_optimization()` function to further reduce these gradients.

@@

We can confirm that the gradients are smaller.

@@

View the confidence intervals on both fixed and random (and variance) parameter estimates using `tidy()`:

@@

`sigma_O` is the spatial random field SD, `sigma_E` is the spatiotemporal random field SD, `sigma_Z` is the spatially varying coefficient random field SD, and `sigma_G` is the random intercept SD.

@@

We can inspect randomized quantile residuals (Fig.\ \ref{fig:qqnorm-owls}), although this method has known issues when based on random effects derived with the Laplace approximation [@thygesen2017a].

@@

More accurate residuals can be generated using MCMC and \pkg{tmbstan}, but this a slower process.
These are randomized quantile residuals where the fixed effects are fixed at their MLE (maximum likelihoood estimate) values and the random effects are sampled with MCMC via \pkg{tmbstan}/\pkg{Stan}.
To calculate these, we have to use a version of our model fit without REML, which we can quickly obtain with the `update()` method.
We would ideally sample with more MCMC samples than this and assess convergence by setting `print_stan_model = TRUE`.
For the purposes of this building this appendix quickly, we will not run this code.

```{r stan-resids-owls, eval=FALSE}
fit_ml <- update(fit_owl, reml = FALSE)
mcmc_resids <- residuals(fit_owl, type = "mle-mcmc", mcmc_iter = 101, mcmc_warmup = 100)
```

We can test for zero-inflation using the `simulate.sdmTMB()` function.
If our data were zero-inflated or overdispersed, we might have needed to use a delta ("hurdle") model instead (e.g., `delta_truncated_nbinom2()` or possibly even `delta_gamma()` if we were willing to treat the counts as continuous.

@@

We can more formally test this with functions from the \pkg{DHARMa} package.

@@

```{r plot-all-effects-owls, out.width="95%", fig.asp=0.8, echo=FALSE, eval=TRUE, dpi=300, fig.cap="Predicted winter owl counts that incorporate all fixed effects and random effects.", fig.pos='h', fig.align='center'}
ggplot(data = p_proj) +
  geom_point(data = p, aes(X, Y, colour = exp(est)), size = 0.4, alpha = 0.5) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    trans = "sqrt", # makes variation in low numbers visible
    # can also trim the extreme high values
    limits = c(0, quantile(exp(p$est), 0.995)), na.value = "yellow",
    guide = guide_colourbar(title.position = "top")
  ) +
  labs(colour = "Predicted counts") +
  theme(axis.text = element_blank(), 
  legend.title = element_text(size = 8, hjust = 0), legend.key.height = unit(0.25, "cm"), 
  axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

```{r plot-spatiotemporal-effects-owls, out.width="95%", fig.asp=0.8, echo=FALSE, eval=TRUE, dpi=300, fig.cap="Spatiotemporal random effects.", fig.pos='h', fig.align='center'}
ggplot(data = filter(p_proj, year > 1978)) +
  geom_point(
    data = filter(p, year > 1978), aes(X, Y, colour = epsilon_st),
    size = 0.4, alpha = 0.5
  ) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_gradient2(guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), 
  legend.title = element_text(size = 8, hjust = 0), legend.key.height = unit(0.25, "cm"), 
  axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

```{r plot-fixed-effects-owls, out.width="95%", fig.asp=0.8, echo=FALSE, eval=TRUE, dpi=300, fig.cap="Predictions for each year based on only the non-random field elements.", fig.pos='h', fig.align='center'}
ggplot(data = filter(p_proj, year > 1978)) +
  geom_point(
    data = filter(p, year > 1978), aes(X, Y, colour = exp(est_non_rf)),
    size = 0.4, alpha = 0.5
  ) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    breaks = c(0.03, 0.05, 0.07),
    guide = guide_colourbar(title.position = "top")
  ) +
  theme(axis.text = element_blank(), 
  legend.title = element_text(size = 8, hjust = 0), legend.key.height = unit(0.25, "cm"), 
  axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

### 2022-11-04 doc/paper-jss/timing.Rmd
Edits

```{r sdmTMBfit-timing, echo=TRUE}
fit_sdmTMB <- sdmTMB(
  observed ~ a1,
  data = sim_dat, mesh = mesh_sdmTMB, family = gaussian(),
  priors = sdmTMBpriors(matern_s = pc_matern(range_gt = 0.05, sigma_lt = 2))
)
```

>>>>

```{r sdmTMBfit-timing, echo=TRUE}
fit_sdmTMB <- sdmTMB(
  observed ~ a1,
  data = sim_dat, 
  mesh = mesh_sdmTMB, 
  family = gaussian(),
  priors = sdmTMBpriors(
    matern_s = pc_matern(range_gt = 0.05, sigma_lt = 2)
  )
)
```
